% Chapter 5 — Conclusion & Future Work

\chapter{Conclusion \& Future Work}
\label{Chapter5}
\lhead{Chapter 5. \emph{Conclusion \& Future Work}}

%========================================================================================
\section{Conclusion}

This thesis has presented \textbf{SkillBridge} — a full-stack, AI-powered freelance
marketplace platform developed as the Final Year Project of three BS Computer Science
students at GC University Lahore:

\begin{center}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{>{\bfseries}l l l}
\toprule
Name & Roll Number & Email \\
\midrule
Muhammad Ali           & 0017-BSCS-22 & muhammadalikhurram01@gmail.com \\
Muhammad Daniyal Tallat& 0001-BSCS-22 & daniyaltallat0@gmail.com \\
Ahmed Raza Ajmal       & 0053-BSCS-22 & ahmedrazaajmal56@gmail.com \\
\bottomrule
\end{tabular}
\end{center}

\vspace{1em}
The project addressed a clearly defined and empirically validated gap in existing
freelance platforms, making three primary research-backed contributions:

\begin{infobox}[Contribution 1 — AI-Powered Project Scoping]
Integration of the Google Gemini LLM into a structured conversational scoping assistant
enables clients — including non-technical stakeholders — to define unambiguous project
requirements through natural language. The AI returns structured JSON fields used to
pre-fill the project posting form, reducing posting time by approximately 40\%.
\end{infobox}

\vspace{0.5em}
\begin{infobox}[Contribution 2 — SkillToken Economy]
A transparent, auditable virtual token system regulates proposal behaviour by associating
a token cost with proposal submission. The immutable \texttt{TokenTransaction} ledger
with six distinct reason codes provides administrators with full auditability of every
token state change, incentivising quality while deterring spam.
\end{infobox}

\vspace{0.5em}
\begin{infobox}[Contribution 3 — Blind Review System]
A simultaneous-reveal review mechanism — where reviews remain sealed until both parties
submit or a 14-day timeout elapses — eliminates the reciprocal rating inflation that
plagues existing platforms, yielding a more honest and informative reputation system.
\end{infobox}

\vspace{0.8em}
Beyond these innovations, SkillBridge delivers a production-quality feature set:
milestone-based contracts with escrow logic, real-time messaging, dispute resolution,
invitation management, freelancer and project browsing, and role-specific dashboards
for Clients, Freelancers, and Administrators.

All \textbf{85 functional test cases} passed successfully. Average API response times
remained under 200\,ms for all locally-served endpoints under a simulated load of 50
concurrent users.

%========================================================================================
\section{Limitations}

\begin{table}[h]
\centering
\caption{Known Limitations and Mitigations}
\label{tab:limitations}
\renewcommand{\arraystretch}{1.5}
\begin{tabularx}{\textwidth}{|p{0.5cm}|p{5cm}|X|}
\hline
\rowcolor{sblight}
\textbf{\#} & \textbf{Limitation} & \textbf{Planned Mitigation} \\ \hline
1 & Payment gateway not integrated (escrow is simulated) & Integrate Stripe Connect with KYC \\ \hline
2 & AI (Gemini) adds 1.8--3.2s latency & Stream responses; cache common queries \\ \hline
3 & Browse uses filters only, no ML recommendations & Train collaborative filtering on BrowseInteraction data \\ \hline
4 & Primarily desktop-focused UI & Develop React Native mobile application \\ \hline
\end{tabularx}
\end{table}

%========================================================================================
\section{Future Work}

\subsection{Payment Gateway Integration}

The highest-priority enhancement is integrating \textbf{Stripe Connect}, which supports
marketplace KYC, the ``charge-and-transfer'' escrow model, and international payouts.
Implementation requires creating Connected Accounts during freelancer onboarding, replacing
simulated payment records with real \texttt{PaymentIntent} and \texttt{Transfer} API
calls, and deploying a Stripe webhook listener.

\subsection{Machine Learning Recommendation Engine}

The \texttt{BrowseInteraction} model already captures the user-project interaction data
required for collaborative filtering. A planned matrix factorisation pipeline (ALS or SVD)
applied to this data would produce personalised project and freelancer recommendations.

\subsection{Mobile Application}

A React Native application sharing API clients with the web frontend would significantly
expand the platform's reach without requiring any backend changes.

\subsection{AI-Enhanced Proposal Shortlisting}

Future iterations could apply Gemini to automatically rank proposals against project
requirements, providing clients with a structured, AI-generated shortlist to reduce
screening time.

\subsection{Internationalisation}

Support for multiple UI languages via \texttt{i18next} would open the platform to South
and Southeast Asian freelancing markets, where English proficiency varies.

%========================================================================================
\section{Final Remarks}

\begin{featurebox}
SkillBridge demonstrates that thoughtfully applied artificial intelligence, economic
mechanism design (token economy), and established behavioural economics principles
(blind reviews) can — in combination — substantially improve the fairness, efficiency,
and trust of digital freelance marketplaces. This project is submitted with confidence
that it meets and exceeds the expectations set in the original FYP proposal.
\end{featurebox}
