% Chapter 2 — Literature Review

\chapter{Literature Review}
\label{Chapter2}
\lhead{Chapter 2. \emph{Literature Review}}

%----------------------------------------------------------------------------------------
\section{Introduction}

This chapter reviews the existing body of knowledge relevant to the SkillBridge platform,
covering: \textit{(i)} digital freelance marketplace platforms, \textit{(ii)} the role of
Artificial Intelligence in hiring tools, \textit{(iii)} token and gamification economies,
and \textit{(iv)} trust, reputation, and blind review mechanisms.

%----------------------------------------------------------------------------------------
\section{Existing Freelance Marketplace Platforms}

\subsection{Upwork}

Upwork is the world's largest freelance platform, connecting over 18 million registered
freelancers with clients in 180+ countries~\citep{Upwork2023}. It supports hourly
contracts (monitored via Work Diary screenshots) and fixed-price milestones. Despite its
scale, Upwork is criticised for its high service fees (up to 20\%), an opaque Job Success
Score algorithm, and the ``Connects'' credit system which disproportionately affects new
freelancers.

\subsection{Fiverr}

Fiverr operates on a Gig-based model where freelancers define pre-packaged services at fixed
price points~\citep{Fiverr2023}. Its strength is low-friction purchasing, but this model
commoditises professional services. The 20\% platform fee and the absence of AI-assisted
project definition are notable gaps.

\subsection{Freelancer.com}

Freelancer.com pioneered competitive bidding. While this promotes market-driven pricing, it
generates significant volumes of low-quality, template-driven proposals and has weak dispute
and credential-verification mechanisms.

\subsection{Comparative Analysis}

\begin{table}[h]
\centering
\caption{Comparative Analysis of Freelance Platforms}
\label{tab:platform_comparison}
\renewcommand{\arraystretch}{1.5}
\begin{tabularx}{\textwidth}{|>{\bfseries}p{3.5cm}|p{2cm}|p{2cm}|p{2cm}|>{\bfseries\color{sbblue}}p{2.5cm}|}
\hline
\rowcolor{sblight}
\textbf{Feature} & \textbf{Upwork} & \textbf{Fiverr} & \textbf{Freelancer.com} & \textbf{SkillBridge} \\ \hline
AI Project Scoping       & \texttimes & \texttimes & \texttimes & \checkmark\ Gemini \\ \hline
Token Economy            & Partial    & \texttimes & \texttimes & \checkmark \\ \hline
Blind Review System      & \texttimes & \texttimes & \texttimes & \checkmark \\ \hline
Milestone Contracts      & \checkmark & \checkmark & \checkmark & \checkmark \\ \hline
Real-time Chat           & \checkmark & \checkmark & \checkmark & \checkmark \\ \hline
Platform Fee             & up to 20\% & 20\%       & 10--20\%   & Configurable \\ \hline
Dispute Resolution       & Basic      & Basic      & Basic      & Admin-managed \\ \hline
Gig / Package Support    & \texttimes & \checkmark & \texttimes & \checkmark \\ \hline
\end{tabularx}
\end{table}

%----------------------------------------------------------------------------------------
\section{Artificial Intelligence in Hiring Platforms}

\subsection{AI-Assisted Matching and Recommendation}

Machine learning has been applied to talent platforms extensively.
\citeauthor{Kenthapadi2017}~(\citeyear{Kenthapadi2017}) described LinkedIn's use of ML for
job-candidate matching, showing that AI-driven recommendations significantly improve
engagement. \citeauthor{Almalis2015}~(\citeyear{Almalis2015}) proposed content-based
filtering using semantic analysis of job descriptions and candidate profiles.

\subsection{Large Language Models for Requirement Engineering}

\begin{infobox}[Key Insight — LLMs for Project Scoping]
Large Language Models (LLMs) can assist non-technical stakeholders in articulating
software requirements through guided conversational prompts, significantly reducing
ambiguity in project definitions~\citep{Ahmed2023}.
\end{infobox}

SkillBridge directly operationalises this insight: the AI scoping module sends a structured
system prompt to the Google Gemini API, instructing it to return a JSON object with
\textit{category}, \textit{subCategory}, \textit{size}, \textit{budgetRange},
\textit{requiredSkills}, and \textit{scopeSummary}.

\subsection{AI for Fraud and Quality Control}

\citeauthor{Rawal2021}~(\citeyear{Rawal2021}) explored ML-based detection of fraudulent
profiles on freelance platforms. SkillBridge addresses this preventively through the
SkillToken economy rather than post-hoc detection.

%----------------------------------------------------------------------------------------
\section{Token Economies and Gamification}

\subsection{Token-Based Incentive Systems}

Token economies use virtual currency to incentivise desirable behaviours and penalise
undesirable ones~\citep{Skinner1938}. In digital platforms, Upwork's ``Connects'' system
is a direct precedent, although it is criticised for opacity.

SkillBridge's SkillToken economy improves on this by providing a \textbf{transparent,
auditable ledger} (\texttt{TokenTransaction} model) with descriptive reason codes, partial
refunds on withdrawal, and admin-controlled adjustments — creating a richer behavioural
signal than Connects.

\subsection{Gamification in Digital Platforms}

\citeauthor{Deterding2011}~(\citeyear{Deterding2011}) defined gamification as ``the use of
game design elements in non-game contexts.'' Profile completion tracking, skill badges, and
reputation scores are gamification elements that SkillBridge implements natively through the
\texttt{profileCompletion} field and the SkillToken balance display in the Freelancer
dashboard.

%----------------------------------------------------------------------------------------
\section{Trust, Reputation, and Blind Review Systems}

\subsection{Online Reputation Systems}

Reputation reduces information asymmetry in online markets~\citep{Resnick2000,
DellarocasStar2003}. A persistent problem is \textit{reciprocal rating inflation} — both
parties give high ratings to avoid retaliation~\citep{Fradkin2015}.

\subsection{Blind Review Mechanisms}

\citeauthor{Fradkin2015}~(\citeyear{Fradkin2015}) empirically demonstrated that Airbnb's
simultaneous-reveal review system significantly improved review honesty. SkillBridge adopts
this pattern: the \texttt{Review.isRevealed} flag remains \texttt{false} until both parties
submit, or a 14-day window elapses.

%----------------------------------------------------------------------------------------
\section{Summary}

\begin{featurebox}
The literature review confirms a clear market gap: no existing commercial platform
integrates \textbf{AI-assisted project scoping}, a \textbf{transparent token economy},
and a \textbf{blind review system} simultaneously. SkillBridge addresses this gap by
synthesising academic research and industry practice into a unified, modern platform.
\end{featurebox}
