%============================================================
%  Chapter 3 – System Requirements
%============================================================
\chapter{System Requirements}

\section{Overview}

This chapter presents the complete set of functional and non-functional requirements for the \projectname{} system. Requirements were elicited through analysis of the problem domain, study of competing platforms, and iterative design sessions.

\section{Stakeholders}

\begin{table}[H]
\centering
\caption{System Stakeholders}
\begin{tabularx}{\textwidth}{lX}
\toprule
\textbf{Stakeholder} & \textbf{Description} \\
\midrule
Client    & A registered user who posts projects, reviews proposals, manages contracts, and leaves reviews. \\
Freelancer & A registered user who submits proposals, performs contracted work, and receives payment. \\
Admin     & A platform operator who moderates content, resolves disputes, and manages platform settings. \\
System    & Automated processes including cron jobs, email notifications, and the AI assistant. \\
\bottomrule
\end{tabularx}
\end{table}

\section{Functional Requirements}

\subsection{Authentication and User Management}

\begin{enumerate}[label=FR-AUTH-\arabic*]
  \item The system shall allow users to register with an email address and password.
  \item The system shall send a One-Time Password (OTP) to the user's email upon registration for verification.
  \item The system shall allow users to log in using verified email and password credentials.
  \item The system shall support Google OAuth 2.0 for social login.
  \item The system shall issue JWT access tokens (short-lived) and refresh tokens (long-lived) upon authentication.
  \item The system shall rotate refresh tokens on each use, invalidating the previous token.
  \item The system shall allow users to reset their password via an email-based secure token link.
  \item The system shall allow users to select a role (CLIENT or FREELANCER) after initial registration.
  \item The system shall enforce role-based access control on all protected routes.
  \item The system shall allow admins to ban users and provide a ban reason.
  \item The system shall allow users to log out, invalidating their current session.
\end{enumerate}

\subsection{Client Features}

\begin{enumerate}[label=FR-CLIENT-\arabic*]
  \item The system shall allow clients to create, edit, save as draft, and publish project listings.
  \item A project listing shall include: title, description, category/subcategory, budget (fixed or hourly), size, deadline, required skills, experience level, language preference, location preference, hiring method, and optional attachments.
  \item The system shall allow clients to view all proposals submitted for their projects.
  \item The system shall allow clients to shortlist, accept, or reject proposals.
  \item The system shall allow clients to browse freelancer profiles with filtering by skills, experience level, rating, availability, and location.
  \item The system shall allow clients to invite specific freelancers to their projects.
  \item The system shall allow clients to manage active contracts, view milestones, and approve or request revision of submitted milestone work.
  \item The system shall allow clients to open a dispute for any active contract.
  \item The system shall allow clients to leave a review after contract completion.
  \item The system shall allow clients to access an AI assistant for project scoping.
  \item The system shall allow clients to manage their profile (company, industry, bio, location, preferences).
\end{enumerate}

\subsection{Freelancer Features}

\begin{enumerate}[label=FR-FRLNC-\arabic*]
  \item The system shall provide a multi-step onboarding flow for freelancers to complete their profile.
  \item Onboarding steps shall include: basic info, skills, portfolio, certificates, education, availability, social links, and verification document upload.
  \item The system shall calculate a profile completion percentage as a ratio of filled mandatory fields.
  \item The system shall allow freelancers to save an onboarding draft without completing all mandatory fields.
  \item The system shall allow freelancers to browse open project listings with filtering and search.
  \item The system shall allow freelancers to save/bookmark projects of interest.
  \item The system shall allow freelancers to submit a proposal (cover letter, price, delivery time) to a project, deducting SkillTokens.
  \item The system shall allow freelancers to withdraw a pending proposal, refunding SkillTokens.
  \item The system shall allow freelancers to view all active contracts and submit milestone deliveries.
  \item The system shall allow freelancers to view their SkillToken balance and full transaction history.
  \item The system shall allow freelancers to create Gig listings (pre-packaged services).
  \item The system shall allow freelancers to leave a review after contract completion.
\end{enumerate}

\subsection{Admin Features}

\begin{enumerate}[label=FR-ADMIN-\arabic*]
  \item The system shall provide an admin dashboard with platform analytics.
  \item The system shall allow admins to view, search, and ban/unban users.
  \item The system shall allow admins to moderate project listings (view, flag, remove).
  \item The system shall allow admins to manage disputes: assign, review, resolve with a resolution note and decision.
  \item The system shall allow admins to grant or deduct SkillTokens from any freelancer.
  \item The system shall log all admin actions with the target entity, action type, and timestamp.
\end{enumerate}

\subsection{SkillToken Economy}

\begin{enumerate}[label=FR-TOKEN-\arabic*]
  \item Each new freelancer shall receive a registration bonus in SkillTokens.
  \item Submitting a proposal shall deduct a fixed quantity of SkillTokens from the freelancer's balance.
  \item Withdrawing a pending proposal shall refund the SkillTokens.
  \item A cron job shall award monthly SkillTokens to active freelancers.
  \item Every token transaction shall be recorded in an immutable ledger with the balance snapshot.
\end{enumerate}

\subsection{AI Features}

\begin{enumerate}[label=FR-AI-\arabic*]
  \item The system shall provide an AI chat assistant accessible from the client dashboard.
  \item The client shall be able to input an informal project description in natural language.
  \item The AI assistant shall return a structured response including: recommended category/subcategory, estimated budget range, recommended project size, and a list of suggested skills.
  \item The AI scoping session shall be stored in the database and optionally linked to the resulting project.
\end{enumerate}

\section{Non-Functional Requirements}

\begin{table}[H]
\centering
\caption{Non-Functional Requirements}
\begin{tabularx}{\textwidth}{llX}
\toprule
\textbf{ID} & \textbf{Category} & \textbf{Requirement} \\
\midrule
NFR-01 & Security    & All passwords shall be hashed using bcrypt before storage. \\
NFR-02 & Security    & All API routes (except public) shall require a valid JWT. \\
NFR-03 & Security    & Sensitive environment variables (keys, secrets) shall never be committed to version control. \\
NFR-04 & Performance & API endpoints shall respond within 500ms for standard CRUD operations under normal load. \\
NFR-05 & Performance & Redis shall be used to cache frequently accessed, rarely mutated data. \\
NFR-06 & Scalability & The database design shall support horizontal scaling via MongoDB Atlas sharding. \\
NFR-07 & Usability   & The frontend shall be fully responsive and functional on viewports from 375px to 1920px wide. \\
NFR-08 & Usability   & The frontend shall support both light and dark themes. \\
NFR-09 & Reliability & File uploads shall be stored in Cloudinary CDN, not on the application server filesystem. \\
NFR-10 & Maintainability & All source code shall be written in TypeScript with strict type checking enabled. \\
NFR-11 & Maintainability & The database schema shall be managed via Prisma migrations. \\
NFR-12 & Availability & Transactional emails shall be delivered via Nodemailer with a retry mechanism. \\
\bottomrule
\end{tabularx}
\end{table}

\section{Use Case Summary}

\begin{table}[H]
\centering
\caption{Primary Use Cases}
\begin{tabularx}{\textwidth}{lllX}
\toprule
\textbf{UC-ID} & \textbf{Actor} & \textbf{Use Case} & \textbf{Brief Description} \\
\midrule
UC-01 & Client     & Post Project         & Client creates and publishes a project listing. \\
UC-02 & Client     & Review Proposals     & Client views, shortlists, accepts proposals. \\
UC-03 & Client     & Manage Contract      & Client tracks milestones and approves deliveries. \\
UC-04 & Client     & AI Scope Project     & Client describes project; AI generates spec. \\
UC-05 & Client     & Browse Freelancers   & Client filters and views freelancer profiles. \\
UC-06 & Freelancer & Complete Onboarding  & Freelancer fills multi-step profile form. \\
UC-07 & Freelancer & Submit Proposal      & Freelancer bids on a project, spending tokens. \\
UC-08 & Freelancer & Submit Milestone     & Freelancer submits deliverable with attachments. \\
UC-09 & Freelancer & View Tokens          & Freelancer views balance and transaction history. \\
UC-10 & Admin      & Resolve Dispute      & Admin reviews evidence and issues resolution. \\
UC-11 & Both       & Leave Review         & Both parties submit post-contract blind reviews. \\
UC-12 & Both       & Message via Chat     & Parties communicate through contract chat room. \\
\bottomrule
\end{tabularx}
\end{table}
