%============================================================
%  Chapter 8 – SkillToken Economy
%============================================================
\chapter{SkillToken Economy}

\section{Overview}

The \textbf{SkillToken} is an internal virtual currency native to the \projectname{} platform. It was designed to solve a critical quality problem that plagues open bidding marketplaces: the proliferation of low-effort, mass-submitted proposals. By making proposal submission non-free (but not monetarily costly), the system creates a credible signal of freelancer intent and filters out disengaged bidders.

\section{Design Rationale}

\begin{itemize}
  \item \textbf{Cost without money}: SkillTokens are earned through participation and cannot be purchased with real money. This prevents Pay-to-Win dynamics while still creating scarcity.
  \item \textbf{Reversibility}: Withdrawing a pending proposal refunds the tokens, so a freelancer who changes their mind is not penalised.
  \item \textbf{Progressive cost}: Token cost scales with project budget — larger, higher-value projects cost more tokens to bid on, ensuring freelancers are selective.
  \item \textbf{Reward for activity}: Monthly token awards incentivise platform engagement and profile maintenance.
\end{itemize}

\section{Token Events}

\begin{table}[H]
\centering
\caption{SkillToken Ledger Events}
\begin{tabularx}{\textwidth}{llllX}
\toprule
\textbf{Event} & \textbf{Type} & \textbf{Reason Enum} & \textbf{Amount} & \textbf{Trigger} \\
\midrule
Registration      & CREDIT & \texttt{REGISTRATION\_BONUS}    & 50   & New freelancer account created. \\
Submit proposal   & DEBIT  & \texttt{PROPOSAL\_SUBMITTED}    & Varies & Proposal submitted to a project. \\
Withdraw proposal & CREDIT & \texttt{PROPOSAL\_WITHDRAWN}    & Varies & Pending proposal withdrawn. \\
Monthly award     & CREDIT & \texttt{MONTHLY\_AWARD}         & 20   & Cron job (last day of month). \\
Admin grant       & CREDIT & \texttt{ADMIN\_GRANT}           & Any  & Admin manually grants tokens. \\
Admin deduct      & DEBIT  & \texttt{ADMIN\_DEDUCT}          & Any  & Admin manually deducts tokens. \\
\bottomrule
\end{tabularx}
\end{table}

\section{Token Cost Calculation}

The \texttt{tokenCalculator.ts} utility determines the proposal cost based on the project's budget:

\begin{lstlisting}[language=TypeScript, caption={Token Cost Calculation Logic}]
export function calculateTokenCost(projectBudget: number): number {
  if (projectBudget < 100)   return 2;
  if (projectBudget < 500)   return 5;
  if (projectBudget < 2000)  return 10;
  if (projectBudget < 10000) return 20;
  return 30; // Premium projects
}
\end{lstlisting}

This tiered structure ensures that a freelancer proposing on a \$50 micro-project pays 2 tokens, while a proposal on a \$15,000 enterprise project costs 30 tokens, proportional to the value at stake.

\section{Token Ledger Model}

Every token movement is recorded as an immutable \texttt{TokenTransaction} document:

\begin{table}[H]
\centering
\caption{TokenTransaction Fields}
\begin{tabularx}{\textwidth}{llX}
\toprule
\textbf{Field} & \textbf{Type} & \textbf{Description} \\
\midrule
\texttt{id}                  & ObjectId               & Primary key. \\
\texttt{freelancerProfileId} & ObjectId               & Links to the freelancer's profile. \\
\texttt{type}                & TokenWalletTxType      & \texttt{CREDIT} or \texttt{DEBIT}. \\
\texttt{reason}              & TokenTxReason          & Reason enum (6 values). \\
\texttt{amount}              & Int                    & Always positive; direction determined by \texttt{type}. \\
\texttt{balanceAfter}        & Int                    & Running balance snapshot after this transaction. \\
\texttt{description}         & String?                & Human-readable note (e.g., project title). \\
\texttt{relatedProposalId}   & ObjectId?              & Linked if reason is proposal-related. \\
\texttt{createdAt}           & DateTime               & Immutable timestamp. \\
\bottomrule
\end{tabularx}
\end{table}

The \texttt{balanceAfter} field enables both fast current-balance lookup (read the most recent transaction) and full historical reconstruction without summing all rows.

\section{Frontend Token Page}

The \texttt{FreelancerTokens.tsx} page provides:

\begin{itemize}
  \item A prominent \textbf{balance card} showing the current token balance.
  \item An \textbf{information section} explaining the token economy (how to earn, how to spend, and the cost tiers).
  \item A \textbf{transaction history table} showing all credits and debits, sorted by most recent, with colour-coded amounts (green for credit, red for debit).
  \item A breakdown of the token cost by proposal value tier.
\end{itemize}

\section{Admin Token Controls}

Admins can grant or deduct tokens from any freelancer via the User Management panel. This provides an escape hatch for:

\begin{itemize}
  \item Compensating freelancers affected by platform errors.
  \item Penalising accounts that abused the system.
  \item Promotional campaigns.
\end{itemize}

All admin token operations generate the corresponding \texttt{ADMIN\_GRANT} or \texttt{ADMIN\_DEDUCT} \texttt{TokenTransaction} and are also logged in \texttt{AdminLog} for audit purposes.
