%============================================================
%  Chapter 11 – Testing and Validation
%============================================================
\chapter{Testing and Validation}

\section{Overview}

This chapter describes the testing strategy employed during the development of \projectname{}. Due to the nature of this project as an academic prototype, the primary focus is on \textbf{integration-level testing} of the API, \textbf{functional validation} of the frontend, and \textbf{data seeding} for end-to-end scenario verification.

\section{Testing Strategy}

\begin{table}[H]
\centering
\caption{Testing Strategy Overview}
\begin{tabularx}{\textwidth}{llX}
\toprule
\textbf{Level} & \textbf{Method} & \textbf{Scope} \\
\midrule
API Integration  & Manual (Postman / HTTP client) & All REST endpoints and auth flows. \\
Data Seeding     & TypeScript seed scripts        & Database population for test scenarios. \\
Frontend         & Manual browser walkthrough     & All page flows, role-specific routes, form validation. \\
Schema           & Prisma type checking           & All model relationships and field types verified at compile time. \\
\bottomrule
\end{tabularx}
\end{table}

\section{API Testing}

Each API endpoint was tested manually using an HTTP client (Postman / cURL) during development. The following categories of tests were performed:

\subsection{Authentication Flow Tests}

\begin{table}[H]
\centering
\caption{Authentication API Test Cases}
\begin{tabularx}{\textwidth}{lXl}
\toprule
\textbf{Test Case} & \textbf{Description} & \textbf{Result} \\
\midrule
TC-AUTH-01 & Register with valid email and password; OTP sent to email.           & Pass \\
TC-AUTH-02 & Register with duplicate email; expect 409 error.                    & Pass \\
TC-AUTH-03 & Verify OTP with correct code within 10 min; access token returned.  & Pass \\
TC-AUTH-04 & Verify OTP with expired code; expect 400 error.                     & Pass \\
TC-AUTH-05 & Login with correct credentials; JWT tokens issued.                  & Pass \\
TC-AUTH-06 & Login with wrong password; expect 401 error.                        & Pass \\
TC-AUTH-07 & Login with banned account; expect 403 with ban reason.              & Pass \\
TC-AUTH-08 & Refresh token endpoint; new access token issued; old cookie replaced. & Pass \\
TC-AUTH-09 & Logout; session deleted; next refresh fails.                        & Pass \\
TC-AUTH-10 & Password reset flow: request email → receive link → set new password. & Pass \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{Project and Proposal Tests}

\begin{table}[H]
\centering
\caption{Project and Proposal Test Cases}
\begin{tabularx}{\textwidth}{lXl}
\toprule
\textbf{Test Case} & \textbf{Description} & \textbf{Result} \\
\midrule
TC-PROJ-01 & Create project as CLIENT role; status = DRAFT.                       & Pass \\
TC-PROJ-02 & Attempt project creation as FREELANCER role; expect 403.             & Pass \\
TC-PROJ-03 & Publish project (DRAFT → OPEN).                                      & Pass \\
TC-PROP-01 & Submit proposal as FREELANCER; token deducted; proposal created.     & Pass \\
TC-PROP-02 & Submit proposal with insufficient balance; expect 400 error.         & Pass \\
TC-PROP-03 & Submit duplicate proposal (same project); expect 409 error.          & Pass \\
TC-PROP-04 & Withdraw pending proposal; tokens refunded.                          & Pass \\
TC-PROP-05 & Accept proposal as CLIENT; Contract created; project = IN\_PROGRESS. & Pass \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{Token Economy Tests}

\begin{table}[H]
\centering
\caption{SkillToken Economy Test Cases}
\begin{tabularx}{\textwidth}{lXl}
\toprule
\textbf{Test Case} & \textbf{Description} & \textbf{Result} \\
\midrule
TC-TOK-01 & New freelancer receives registration bonus of 50 tokens.            & Pass \\
TC-TOK-02 & Token transaction ledger records correct \texttt{balanceAfter}.     & Pass \\
TC-TOK-03 & Admin grants tokens to freelancer; balance and ledger updated.      & Pass \\
TC-TOK-04 & Admin deducts tokens from freelancer; balance and ledger updated.   & Pass \\
\bottomrule
\end{tabularx}
\end{table}

\section{Data Seeding}

The \texttt{prisma/} directory contains a suite of TypeScript seed scripts used to populate the database with realistic test data:

\begin{table}[H]
\centering
\caption{Database Seed Scripts}
\begin{tabularx}{\textwidth}{lX}
\toprule
\textbf{File} & \textbf{Content} \\
\midrule
\texttt{seed.ts}            & Master seed orchestrator; calls all other seed functions in correct order. \\
\texttt{seed\_metadata.ts}  & Populates the skills catalogue, categories, subcategories, languages, and locations. \\
\texttt{seedFreelancers.ts} & Creates 50 realistic freelancer accounts with complete profiles, skills, portfolio items, and token balances. \\
\texttt{seedClients.ts}     & Creates 20 client accounts with company profiles and posted projects in various statuses. \\
\end{tabularx}
\end{table}

Seeds are executed with:

\begin{lstlisting}[language=bash, caption={Running the Database Seed}]
npm run seed
# or
npx tsx prisma/seed.ts
\end{lstlisting}

The seed data provides a fully populated database for development and demonstration purposes, including freelancers with varying SkillToken balances, clients with open projects, proposals in multiple states, and active contracts with milestones.

\section{Frontend Validation Testing}

All frontend pages were manually tested in both light and dark themes across viewport widths of 375px (mobile), 768px (tablet), and 1920px (desktop). The following key flows were validated:

\begin{enumerate}
  \item \textbf{Full registration-to-dashboard flow}: Signup → OTP verification → Role selection → Dashboard.
  \item \textbf{Freelancer onboarding}: All 8 steps of the onboarding wizard, including image uploads and draft saving.
  \item \textbf{Project posting}: All 8 steps of the PostProject form, including draft save and publish.
  \item \textbf{Proposal submission}: Browse projects → View project → Submit proposal → Verify token deducted.
  \item \textbf{Contract workflow}: Accept proposal → View contract → Submit milestone → Approve milestone.
  \item \textbf{AI scoping}: Enter project description → Receive structured output → Pre-fill PostProject form.
  \item \textbf{Admin flows}: Login as admin → Ban user → Resolve dispute → Grant tokens.
\end{enumerate}

\section{TypeScript Type Safety}

TypeScript's static type system provided a continuous layer of correctness checking throughout development. The Prisma-generated client types ensured that all database queries and mutations were type-correct at compile time, preventing entire categories of runtime errors that are common in untyped JavaScript backends. Running \texttt{npx tsc --noEmit} reported zero type errors in the final submission codebase.

\section{Known Limitations}

\begin{itemize}
  \item No automated unit tests or integration tests are implemented (planned future work).
  \item Payment gateway integration is simulated via status fields; no real money flows are tested.
  \item Real-time messaging (Socket.IO) is architecturally planned but not fully implemented in the current version — chat rooms and messages are persisted but live push is handled via polling.
  \item OTP rate-limiting is not yet implemented, leaving the OTP endpoint theoretically vulnerable to brute-force within the 10-minute window.
\end{itemize}

