%============================================================
%  Appendix A – API Endpoint Reference
%============================================================
\chapter{API Endpoint Reference}
\label{app:api}

This appendix documents all REST API endpoints exposed by the \projectname{} backend. Protected routes require a \texttt{Bearer} JWT access token in the \texttt{Authorization} header unless stated otherwise.

\section*{Legend}

\begin{itemize}
  \item \textbf{[PUB]} — Public (no authentication required)
  \item \textbf{[AUTH]} — Requires authentication (any role)
  \item \textbf{[CLIENT]} — Requires CLIENT role
  \item \textbf{[FRLNC]} — Requires FREELANCER role
  \item \textbf{[ADMIN]} — Requires ADMIN role
\end{itemize}

%------------------------------------------------------------
\section*{Authentication — \texttt{/api/auth}}

\begin{longtable}{p{1.5cm} p{5.5cm} p{1.5cm} p{5.5cm}}
\toprule
\textbf{Method} & \textbf{Endpoint} & \textbf{Access} & \textbf{Description} \\
\midrule
\endhead
\bottomrule
\endfoot
POST & \texttt{/api/auth/signup}            & PUB    & Register new user account; triggers OTP email. \\
POST & \texttt{/api/auth/verify-otp}        & PUB    & Verify email with OTP code. \\
POST & \texttt{/api/auth/login}             & PUB    & Login with email \& password; returns JWT tokens. \\
POST & \texttt{/api/auth/refresh}           & PUB    & Rotate refresh token; return new access token. \\
POST & \texttt{/api/auth/logout}            & AUTH   & Invalidate current session. \\
POST & \texttt{/api/auth/forgot-password}   & PUB    & Send password reset email link. \\
POST & \texttt{/api/auth/reset-password}    & PUB    & Reset password using the emailed token. \\
GET  & \texttt{/api/auth/google}            & PUB    & Initiate Google OAuth 2.0 flow. \\
GET  & \texttt{/api/auth/google/callback}   & PUB    & Google OAuth 2.0 callback handler. \\
POST & \texttt{/api/auth/complete-google-signup} & AUTH & Set role after Google sign-in. \\
\end{longtable}

%------------------------------------------------------------
\section*{Projects — \texttt{/api/projects}}

\begin{longtable}{p{1.5cm} p{5.5cm} p{1.7cm} p{5.3cm}}
\toprule
\textbf{Method} & \textbf{Endpoint} & \textbf{Access} & \textbf{Description} \\
\midrule
\endhead
\bottomrule
\endfoot
GET    & \texttt{/api/projects}               & PUB    & Get all open projects (paginated, filtered). \\
POST   & \texttt{/api/projects}               & CLIENT & Create a new project (status: DRAFT). \\
GET    & \texttt{/api/projects/:id}            & PUB    & Get project by ID. \\
PATCH  & \texttt{/api/projects/:id}            & CLIENT & Update project details. \\
DELETE & \texttt{/api/projects/:id}            & CLIENT & Delete a project. \\
PATCH  & \texttt{/api/projects/:id/publish}    & CLIENT & Transition project DRAFT → OPEN. \\
PATCH  & \texttt{/api/projects/:id/status}     & CLIENT & Update project status (e.g., COMPLETED). \\
\end{longtable}

%------------------------------------------------------------
\section*{Proposals — \texttt{/api/proposals}}

\begin{longtable}{p{1.5cm} p{6cm} p{1.5cm} p{5cm}}
\toprule
\textbf{Method} & \textbf{Endpoint} & \textbf{Access} & \textbf{Description} \\
\midrule
\endhead
\bottomrule
\endfoot
POST   & \texttt{/api/proposals}                          & FRLNC  & Submit proposal; deducts SkillTokens. \\
GET    & \texttt{/api/proposals/my}                        & FRLNC  & Get all proposals submitted by current freelancer. \\
GET    & \texttt{/api/proposals/project/:projectId}        & CLIENT & Get all proposals for a specific project. \\
PATCH  & \texttt{/api/proposals/:id/withdraw}              & FRLNC  & Withdraw pending proposal; refunds tokens. \\
PATCH  & \texttt{/api/proposals/:id/shortlist}             & CLIENT & Mark proposal as shortlisted. \\
PATCH  & \texttt{/api/proposals/:id/accept}                & CLIENT & Accept proposal; creates Contract. \\
PATCH  & \texttt{/api/proposals/:id/reject}                & CLIENT & Reject proposal. \\
\end{longtable}

%------------------------------------------------------------
\section*{Freelancers — \texttt{/api/freelancers}}

\begin{longtable}{p{1.5cm} p{6.5cm} p{1.5cm} p{4.5cm}}
\toprule
\textbf{Method} & \textbf{Endpoint} & \textbf{Access} & \textbf{Description} \\
\midrule
\endhead
\bottomrule
\endfoot
GET   & \texttt{/api/freelancers/profile}                     & FRLNC  & Get own freelancer profile. \\
PATCH & \texttt{/api/freelancers/onboarding/basic-info}       & FRLNC  & Update basic info (step 1). \\
PATCH & \texttt{/api/freelancers/onboarding/skills}           & FRLNC  & Update skills (step 2). \\
PATCH & \texttt{/api/freelancers/onboarding/portfolio}        & FRLNC  & Update portfolio items (step 3). \\
PATCH & \texttt{/api/freelancers/onboarding/certificates}     & FRLNC  & Update certificates (step 4). \\
PATCH & \texttt{/api/freelancers/onboarding/education}        & FRLNC  & Update education (step 5). \\
PATCH & \texttt{/api/freelancers/onboarding/availability}     & FRLNC  & Update availability \& rates (step 6). \\
PATCH & \texttt{/api/freelancers/onboarding/social-links}     & FRLNC  & Update social links (step 7). \\
PATCH & \texttt{/api/freelancers/onboarding/verification}     & FRLNC  & Upload ID document (step 8). \\
GET   & \texttt{/api/freelancers/:id}                         & AUTH   & Get freelancer public profile by ID. \\
\end{longtable}

%------------------------------------------------------------
\section*{Tokens — \texttt{/api/tokens}}

\begin{longtable}{p{1.5cm} p{5.5cm} p{1.5cm} p{5.5cm}}
\toprule
\textbf{Method} & \textbf{Endpoint} & \textbf{Access} & \textbf{Description} \\
\midrule
\endhead
\bottomrule
\endfoot
GET  & \texttt{/api/tokens/balance}           & FRLNC  & Get current SkillToken balance. \\
GET  & \texttt{/api/tokens/history}           & FRLNC  & Get full token transaction history. \\
POST & \texttt{/api/tokens/admin/grant}       & ADMIN  & Grant tokens to a freelancer. \\
POST & \texttt{/api/tokens/admin/deduct}      & ADMIN  & Deduct tokens from a freelancer. \\
\end{longtable}

%------------------------------------------------------------
\section*{Browse — \texttt{/api/browse}}

\begin{longtable}{p{1.5cm} p{5.5cm} p{1.5cm} p{5.5cm}}
\toprule
\textbf{Method} & \textbf{Endpoint} & \textbf{Access} & \textbf{Description} \\
\midrule
\endhead
\bottomrule
\endfoot
GET  & \texttt{/api/browse/freelancers}  & AUTH & Browse freelancer profiles with filters and scoring. \\
GET  & \texttt{/api/browse/projects}     & AUTH & Browse projects with filters. \\
\end{longtable}

%------------------------------------------------------------
\section*{Tracking — \texttt{/api/track}}

\begin{longtable}{p{1.5cm} p{5.5cm} p{1.5cm} p{5.5cm}}
\toprule
\textbf{Method} & \textbf{Endpoint} & \textbf{Access} & \textbf{Description} \\
\midrule
\endhead
\bottomrule
\endfoot
POST   & \texttt{/api/track/view/:projectId}     & FRLNC & Record a project view interaction. \\
POST   & \texttt{/api/track/save/:projectId}     & FRLNC & Save/bookmark a project. \\
DELETE & \texttt{/api/track/save/:projectId}     & FRLNC & Remove a project bookmark. \\
GET    & \texttt{/api/track/saved}               & FRLNC & Get all saved/bookmarked projects. \\
\end{longtable}

%------------------------------------------------------------
\section*{Other Endpoints}

\begin{longtable}{p{1.5cm} p{5.5cm} p{1.5cm} p{5.5cm}}
\toprule
\textbf{Method} & \textbf{Endpoint} & \textbf{Access} & \textbf{Description} \\
\midrule
\endhead
\bottomrule
\endfoot
GET  & \texttt{/api/categories}          & PUB   & List all categories with subcategories. \\
GET  & \texttt{/api/metadata/languages}  & PUB   & List available languages. \\
GET  & \texttt{/api/metadata/locations}  & PUB   & List available locations. \\
POST & \texttt{/api/ai/scope}            & CLIENT & Submit a project description to the AI scoping assistant. \\
\end{longtable}

